Modeling the impacts of park access on health outcomes: a choice-based accessibility approach

### 2019-03-05 manuscript.Rnw
Incorporate changes from John Taylor

%% Front matter =========================
\title{The impacts of park access on health outcomes: A spatial comprehensive approach}

\author{Nico Boyd\affilnum{1}, Gregory S. Macfarlane\affilnum{2},
John Taylor\affilnum{1}, and Kari Watkins\affilnum{1}}
\runninghead{Boyd, et al.}

@@

Historically, most large-scale health promotion efforts focused on
individual-level interventions intended to educate people about healthy
lifestyles and behaviors, touching on topics including diet and exercise.
According to \citet{Duhl1999} however, the field of public health has
moved toward a new paradigm that is grounded in a social model of health,
treating health ``as an outcome of the effects of socioeconomic status, culture,
environmental conditions, housing, employment and community influences.'' (p. 7)
Further, \citet{Coutts2008} maintains that awareness of the impacts of the built
environment on health will increase connections between the practices of public
health and urban planning. As professionals from both fields begin to operate
from the assumption that the design and configuration of the built environment
can facilitate or inhibit physical activity, they are increasingly looking to
public spaces like parks and greenways as key elements of the built environment
that can support physical activity \citep{Bedimo-Rung2005a}.

@@

We apply this measurement to study the link between park
accessibility and attractiveness and Census tract-level aggregate physical
activity participation and obesity rates in New York City, controlling for
spatial correlation and tract-level socioeconomic characteristics.
We find a positive relationship where the least park-accessible tracts have an
expected physical activity participation rate roughly a full percentage point
lower than the most accessible tracts; a traditional quarter-mile buffer
analysis estimates an similar albeit marginally less significant effect.
However, we also find a strong and significant correlation between
the choice-based park accessibility metric and decreased obesity rates --- again
controlling for spatial correlation, socioeconomic attributes, and physical
activity rates. In this case a traditional buffer analysis does not show
significance.

@@

Note that this formulation is an extension of gravity-based accessibility
statistics, in the same way that the gravity-based trip distribution model
is a specific case of
a destination-choice model \citep{Daly1982}. But the extension is meaningful and
allows us to consider more attributes of the park or the individual beyond size
and impedance. For example, we can aggregate the number of tweets emanating from
a park as a measure of the park's attractiveness beyond its size, and the
utility for the choice maker from Equation \ref{eq:v_sd} then becomes
\begin{equation}\label{eq:v_sdt}
V_{ij} = size_j\lambda_s + distance_{ij}\lambda_d + tweets_j\lambda_t
\end{equation}

@@

\section{Empirical Application}
In this section, we describe an experiment with data for New York City, where
we compute a holistic logsum-based accessibility to parks measure and model the
relationship between this measure and physical activity rates, controlling for
spatial effects and socioeconomic factors. We subsequently model the effect
of the accessibility metric on obesity rates, accounting for physical activity and
the other controls. We then model the relationship between a modified logsum
considering geolocated Twitter data and physical activity and obesity rates.

@@

In a destination choice framework, the tracts represent the ``origins'' and
``destinations'' are parks in New York City.
We retrieved a shapefile of public parks and greenspaces within New York City’s
municipal boundaries and checked it for accuracy and relevance \citep{nycparks}.
Upon inspection, we removed several facilities that do not qualify as
publicly accessible green space, such as Yankee Stadium and its surrounding
parking lots. We also removed parks of less than half an acre in size, as these
appear to be predominantly vacant lots or parking strips rather than legitimate
public green space. This leaves us with \Sexpr{nrow(parks)} parks. It is worth
noting that the particular dataset classifies specific facilities within larger
parks as independent features: baseball fields within Central Park, for example,
are themselves delineated green spaces. We elect to retain these independent
features as they may generate additional physical activity participation aside
from the containing park. Additionally, large cemeteries are not considered
open space in these data, though people may use the space for physical activities
such as walking or running. For each park we calculate its size in acres, and
identify the number of geolocated tweets emanating from the park in September
2014, as a proxy for the park's use or popularity.

@@

Increased physical activity and decreased obesity rates are critical measures of
improvement in public health. Although many have theorized the link between park
accessibility and these metrics, previous literature has been mixed. Using New
York City as a case study, we presented a holistic and flexible measurement for
park accessibility that compares the continuous distance to all parks in the
region, weighted against the size of the park and its other amenities. In terms
of physical activity, we found a positive relationship where the least
park-accessible tracts have an expected physical activity participation rate
roughly one percentage point lower than the most accessible tracts. In terms of
obesity, we also found a strong and significant correlation between increased
park accessibility and decreased obesity rates, controlling for
spatial correlation, socioeconomic attributes, and physical activity rates.

In both cases a traditional, buffer-based analysis estimated a weaker relationship
with greater uncertainty. Buffer analyses are relatively common in the literature,
perhaps because of the widespread availability of GIS software. In spite of their
widespread use, they are relatively limited in terms of both their theoretical
underpinnings and their flexibility to accommodate attributes of parks beyond
their proximity. And even proximity may not be adequately handled, as the buffer
distance may be arbitrarily asserted by the researcher. Adopting choice-based
accessibilities of the kind used in this study will allow researchers to
encompass the full universe of park amenities in their accessibility analyses.
This will in turn enable planners to consider how multiple attributes of a park
--- from its location to its size to its amenities and beyond --- benefit the health
of the community the park serves.

>>>>

%% Front matter =========================
\title{Modeling the impacts of park access on health outcomes: a choice-based accessibility approach}

\author{Nico Boyd\affilnum{1}, Gregory S. Macfarlane\affilnum{2},
John E. Taylor\affilnum{1}, and Kari Watkins\affilnum{1}}
\runninghead{Boyd, et al.}

@@

Historically, most large-scale health promotion efforts focused on
individual-level interventions intended to educate people about healthy
lifestyles and behaviors, touching on topics including diet and exercise.
However, \citet{Duhl1999} claim the field of public health has
moved toward a new paradigm that is grounded in a social model of health,
treating health ``as an outcome of the effects of socioeconomic status, culture,
environmental conditions, housing, employment and community influences.'' (p. 7)
Further, \citet{Coutts2008} maintains that awareness of the impacts of the built
environment on health will increase connections between the practices of public
health and urban planning. As professionals from both fields begin to operate
from the assumption that the design and configuration of the built environment
can facilitate or inhibit physical activity, they are increasingly looking to
public spaces like parks and greenways as key elements of the built environment
that can support physical activity \citep{Bedimo-Rung2005a}.

@@

We apply this measurement to study the link between park
accessibility and attractiveness and Census tract-level aggregate physical
activity participation and obesity rates in New York City, controlling for
spatial correlation and tract-level socioeconomic characteristics.
We find a positive relationship where the least park-accessible tracts have an
expected physical activity participation rate roughly a full percentage point
lower than the most accessible tracts; a traditional quarter-mile buffer
analysis estimates a similar albeit marginally less significant effect.
However, we also find a strong and significant correlation between
the choice-based park accessibility metric and decreased obesity rates --- again
controlling for spatial correlation, socioeconomic attributes, and physical
activity rates. In this case a traditional buffer analysis does not show
significance.

@@

Note that this formulation is an extension of gravity-based accessibility
statistics, in the same way that the gravity-based trip distribution model
is a specific case of
a destination-choice model \citep{Daly1982}. But the extension is meaningful and
allows us to consider more attributes of the park or the individual beyond size
and impedance. For example, we can aggregate the number of tweets emanating from
a park as a measure of the park's attractiveness beyond its size. The
utility for the choice maker from Equation \ref{eq:v_sd} then becomes
\begin{equation}\label{eq:v_sdt}
V_{ij} = size_j\lambda_s + distance_{ij}\lambda_d + tweets_j\lambda_t
\end{equation}

@@

\section{Empirical Application}
In this section, we develop a model with data for New York City, where
we compute a holistic logsum-based accessibility to parks measure and model the
relationship between this measure and physical activity rates, controlling for
spatial effects and socioeconomic factors. We subsequently model the effect
of the accessibility metric on obesity rates, accounting for physical activity and
the other controls. Finally, we model the relationship between a modified logsum
measure considering geolocated Twitter data and physical activity and obesity rates.

@@

In a destination choice framework, the tracts represent the ``origins'' and
``destinations'' are parks in New York City.
We retrieved a shapefile of public parks and greenspaces within New York City’s
municipal boundaries and checked it for accuracy and relevance \citep{nycparks}.
Upon inspection, we removed several facilities that do not qualify as
publicly accessible green space, such as Yankee Stadium and its surrounding
parking lots. We also removed parks of less than half an acre in size, as these
appear to be predominantly vacant lots or parking strips rather than legitimate
public green space. This leaves us with \Sexpr{nrow(parks)} parks. It is worth
noting that the particular dataset classifies specific facilities within larger
parks as independent features: baseball fields within Central Park, for example,
are themselves delineated green spaces. We elect to retain these independent
features as they may generate additional physical activity participation aside
from the containing park. Additionally, large cemeteries are not considered
open space in these data, though people may use the space for physical activities
such as walking or running. For each park we calculate its size in acres. As a
proxy for the park's use or popularity, we add a measure of Twitter activity. Using
a Twitter application programming interface (API) implemented with the Python
package Tweepy, we collected and stored tweets containing geotags of the exact
geographic coordinates \citep{Wang2016}.  From this information, we were
able to segregate tweets that were generated within a park in September 2014.

@@

Increased physical activity and decreased obesity rates are critical measures of
improvement in public health. Although many have theorized the link between park
accessibility and these metrics, previous literature has produced mixed
findings, perhaps owing to the range of variables modeled and the coarse spatial
scale of the analyses. Using New York City as a case study, we presented a
holistic and flexible measurement for
park accessibility that compares the continuous distance to all parks in the
region, weighted against the size of the park and its other amenities. In terms
of physical activity, we found a positive relationship where the least
park-accessible tracts have an expected physical activity participation rate
roughly one percentage point lower than the most accessible tracts. In terms of
obesity, we also found a strong and significant correlation between increased
park accessibility and decreased obesity rates, controlling for
spatial correlation, socioeconomic attributes, and physical activity rates.

In both cases a traditional, buffer-based analysis estimated a weaker relationship
with greater uncertainty. Buffer analyses are relatively common in the literature,
perhaps because of the widespread availability of GIS software. In spite of their
widespread use, they are relatively limited in terms of both their theoretical
underpinnings and their flexibility to accommodate attributes of parks beyond
their proximity. And even proximity may not be adequately handled, as the buffer
distance may be arbitrarily asserted by the researcher. The model we develop and
apply in this paper extends buffer driven models to contribute a more
comprehensive and flexible approach for measuring the impact of park access on
health outcomes. Adopting choice-based accessibilities of the kind used in this
study will allow researchers to
encompass the full range of park amenities in their accessibility analyses.
This will in turn enable planners to consider how multiple attributes of a park
--- from its location to its size to its amenities and beyond --- benefit the health
of the community the park serves.
